---
title: "Southern Elephant Seal Pup Dispersal Analysis"
author: "Dahlia Foo"
date: today
format:
    html:
        embed-resources: true
        code-fold: true
        code-line-numbers: true
---

```{r setup, include = FALSE}
#
knitr::opts_chunk$set(echo = TRUE)
options(width = 10000)

```

```{r load-packages}
#| output: false

library(tidyverse)
library(circular)
library(sf)
library(move)
library(conflicted)
library(patchwork)
library(skimr)
library(gtsummary)
library(gt)
library(gtExtras)
library(summarytools)
library(GGally)
library(broom)
library(DHARMa)
library(MuMIn)
library(here)

conflicts_prefer(dplyr::filter, dplyr::select, dplyr::mutate, dplyr::group_by)

theme_set(theme_bw())

# Define output folder
output_folder <- here("output", "dispersal_analysis_3")

# Create the output folder if it doesn't exist
dir.create(output_folder, showWarnings = FALSE, recursive = TRUE)
```

## Data Loading and Preprocessing

```{r load-data}
#| output: false

seal_data_path <- here("output", "tracks_processed_12h.rds")
female_data_path <- here("data", "adult_female_locs.rds")
particle_data_path <- here("output", "particle-trace-186.13m.rds")
surface_particle_data_path <- here("output", "currently_particleTrace.rds")

source(here("code", "functions", "functions.R"))

colony <- cbind(158.95, -54.5)

# Include your function definitions

calculate_distance_from_start <- function(data, colony = cbind(158.95, -54.5)) {
    data %>%
        group_by(id) %>%
        mutate(
            distance_from_start = geosphere::distGeo(
                cbind(colony[1], colony[2]),
                cbind(lon, lat)
            )
        )
}

identify_outbound_trip <- function(data) {
    data %>%
        calculate_distance_from_start() %>%
        group_by(id) %>%
        arrange(id, date) %>%
        mutate(
            max_distance = max(distance_from_start),
            is_outbound = distance_from_start <= max_distance & row_number() <= which.max(distance_from_start)
        ) %>%
        ungroup()
}

normalize_longitude <- function(lon) {
    ((lon + 180) %% 360) - 180
}

interpolate_track <- function(data) {
    data <- data %>% as.data.frame()

    # Print the number of rows before processing
    cat("Total rows before processing:", nrow(data), "\n")

    processed_data <- data %>%
        group_split(id) %>%
        map_dfr(~ {
            id <- .$id %>% unique()
            cat("Processing ", id, "...\n")
            initial_rows <- nrow(.)
            do_interpolate <- FALSE

            # Create move object
            move_obj <- move(
                x = .$lon,
                y = .$lat,
                time = .$date,
                proj = CRS("+proj=longlat +datum=WGS84"),
                data = .,
            )

            # Only interpolate if there are gaps larger than 1 day
            time_diffs <- difftime(.$date, lag(.$date), units = "days") %>% as.numeric()
            if (all(time_diffs %>% na.omit() != 1)) {
                cat("time interval not 1 day. interpolating...\n")
                do_interpolate <- TRUE
                interpolated <- interpolateTime(move_obj, time = as.difftime(24, units = "hours"), spaceMethod = "greatcircle")
            } else {
                interpolated <- move_obj
            }

            result <- interpolated %>%
                as_tibble() %>%
                mutate(id = id, date = as.Date(date)) %>%
                dplyr::select(-lon, -lat) %>%
                rename(lon = coords.x1, lat = coords.x2) %>%
                dplyr::select(id, date, lon, lat, everything())

            if (do_interpolate) {
                cat("interpolated from", initial_rows, "to", nrow(result), "rows\n")
            }
            return(result)
        })

    return(processed_data)
}

preprocess_track <- function(data) {
    data %>%
        mutate(lon = normalize_longitude(lon)) %>%
        # drop_na(lon, lat) %>%
        arrange(id, date) %>%
        group_by(id) %>%
        interpolate_track() %>%
        calculate_bearings() %>%
        mutate(bearing = make_circular(bearing)) %>%
        identify_outbound_trip()
}

# Calculate days since start for each id
calculate_days_since_start <- function(data) {
    data %>%
        group_by(id) %>%
        arrange(id, date) %>%
        mutate(days_since_start = as.numeric(difftime(date, first(date), units = "days"))) %>%
        ungroup()
}

```

```{r preprocess-data}
#| output: false
locw <- readRDS(seal_data_path) %>%
    filter(trip == 1, SUS == FALSE) %>%
    dplyr::select(-daysFromDeployment, -SUS, -trip, -dist2col, -haulout) %>%
    preprocess_track() %>%
    convert2polarsf(remove_coords = FALSE) %>%
    filter(is_outbound)

locf <- readRDS(female_data_path) %>%
    preprocess_track() %>%
    convert2polarsf(remove_coords = FALSE) %>%
    filter(is_outbound)

locp <- readRDS(particle_data_path) %>%
    preprocess_track() %>%
    convert2polarsf(remove_coords = FALSE)

locps <- readRDS(surface_particle_data_path) %>%
    preprocess_track() %>%
    convert2polarsf(remove_coords = FALSE)
```

## Data Summary

```{r data-summary}
#| echo: false
cat("Number of unique ids in locf:", locf$id %>% unique() %>% length(), "\n")
cat("Number of unique ids in locw:", locw$id %>% unique() %>% length(), "\n")
cat("Number of unique ids in locps:", locps$id %>% unique() %>% length(), "\n")
cat("Number of unique ids in locp:", locp$id %>% unique() %>% length(), "\n")
```

## Plotting All Tracks

```{r plot-all-tracks}
#| fig-cap: "All tracks plot with ORSI fronts"

# Option 1: Colorblind-friendly palette with muted tones
palette1 <- c(
    "Female" = "#999999", # Muted grey
    "Weaner" = "#E69F00", # Orange
    "Particle 0m" = "#56B4E9", # Light blue
    "Particle 186m" = "#0072B2" # Dark blue
)

# Option 2: Palette with richer colors
palette2 <- c(
    "Female" = "#7F7F7F", # Warm grey
    "Weaner" = "#D55E00", # Rich orange-red
    "Particle 0m" = "#69B3E7", # Sky blue
    "Particle 186m" = "#1B4B8C" # Deep navy
)

# Option 3: Palette with more contrast
palette3 <- c(
    "Female" = "#8C8C8C", # Medium grey
    "Weaner" = "#CC3311", # Bright red
    "Particle 0m" = "#33B2FF", # Bright blue
    "Particle 186m" = "#004C99" # Dark blue
)


# Get the bounding box of all tracks combined
bb <- sf::st_bbox(
    sf::st_union(
        sf::st_union(locf),
        sf::st_union(locw),
        sf::st_union(locps),
        sf::st_union(locp)
    )
)

# Add nudge values for the labels
orsi_sf <- orsi_sf %>%
    mutate(
        nudge_x = c(-1000, -4800, -4500, -4000),
        nudge_y = c(0, 5300, 4800, 3000),
        front = toupper(front)
    )

# Convert Macquarie Island coordinates to sf object
mq_sf <- st_as_sf(data.frame(lon = 158.95, lat = -54.5),
    coords = c("lon", "lat"),
    crs = 4326
) # assuming WGS84


all_tracks_plot <- ggplot() +
    geom_sf(data = orsi_sf, linetype = "dashed", colour = "grey30") +
    geom_sf_text(
        data = orsi_sf,
        aes(label = front),
        nudge_x = orsi_sf$nudge_x,
        nudge_y = orsi_sf$nudge_y
    ) +
    geom_sf(data = locf, aes(color = "Female"), alpha = 0.4, size = .5) +
    geom_sf(data = locw, aes(color = "Weaner"), alpha = 0.4, size = .5) +
    geom_sf(data = locps, aes(color = "Particle 0m"), alpha = 0.4, size = .5) +
    geom_sf(data = locp, aes(color = "Particle 186m"), alpha = 0.4, size = .5) +
    geom_sf(data = mq_sf, shape = 17, size = 3, color = "black") +
    theme_bw() +
    # Set the x and y limits using the bounding box
    coord_sf(
        xlim = c(bb["xmin"], bb["xmax"]),
        ylim = c(bb["ymin"], bb["ymax"] + 1000)
    ) +
    scale_color_manual(values = palette3) +
    labs(color = "Group", x = "Longitude", y = "Latitude")

all_tracks_plot
ggsave(file.path(output_folder, "all_tracks_plot.png"), all_tracks_plot, width = 8, height = 4)

```


## Bearing Analysis

```{r bearing-analysis}
#| fig-cap: "All sequential outbound bearings"
#| warning: false

## Create circular plot of all bearings for each data set
plot_bearing_histogram <- function(data, title = NULL) {
    # Calculate mean bearing
    mean_bearing <- mean.circular(data$bearing, na.rm = TRUE)
    # mean_bearing <- data$bearing %>% median.circular(na.rm = TRUE)

    p <- data %>%
        ggplot(aes(x = bearing)) +
        geom_histogram(binwidth = 10, boundary = 0) +
        coord_polar(start = 0, direction = 1) +
        scale_x_continuous(
            breaks = seq(0, 360, by = 45),
            limits = c(0, 360),
            labels = c("N", "", "E", "", "S", "", "W", "", "N")
        ) +
        theme_bw() +
        labs(x = NULL, y = "Count") +
        theme(
            axis.text.y = element_blank(),
            axis.ticks.y = element_blank()
        ) +
        # Add arrow for mean bearing
        annotate("segment",
            x = mean_bearing, y = 0, xend = mean_bearing, yend = Inf,
            arrow = arrow(length = unit(0.5, "cm")), color = "red", linewidth = 1
        ) +
        # Add text label for mean bearing
        annotate("text",
            x = mean_bearing, y = Inf, label = sprintf("%.1f°", as.numeric(mean_bearing)),
            color = "red", vjust = 1.1, hjust = -0.1
        )

    if (!is.null(title)) {
        p <- p + labs(title = title)
    } else {
        p <- p + theme(plot.title = element_blank())
    }

    return(p)
}

# All sequential outbound bearings
p_list <- list(locw, locps, locf, locp)
p_names <- c("Weaner", "Particle 0m", "Female", "Particle 186m")

# Map of all bearings for each data set
all_bearings_plot <- map2(p_list, p_names, plot_bearing_histogram) %>%
    wrap_plots(ncol = 2)

all_bearings_plot

ggsave(file.path(output_folder, "all_bearings_plot.png"), all_bearings_plot, width = 12, height = 10)
```

## Create mean bearings by id

```{r mean-bearings-by-id}
#| fig-cap: "Mean bearings by id"

# Calculate mean bearings by id
process_by_id <- function(data) {
    data %>%
        group_by(id) %>%
        summarize(mean_bearing = mean.circular(bearing, na.rm = TRUE), sd_bearing = sd.circular(bearing, na.rm = TRUE)) %>%
        rename(bearing = mean_bearing)
}


locw_by_id <- locw %>% process_by_id()
locps_by_id <- locps %>% process_by_id()
locf_by_id <- locf %>% process_by_id()
locp_by_id <- locp %>% process_by_id()

p_list <- list(locw_by_id, locps_by_id, locf_by_id, locp_by_id)
p_names <- c("Weaner", "Particle 0m", "Female", "Particle 186m")

# Map of mean bearings by id
mean_bearings_plot <- map2(p_list, p_names, plot_bearing_histogram) %>%
    wrap_plots(ncol = 2, guides = "collect")

mean_bearings_plot

ggsave(file.path(output_folder, "mean_bearings_plot.png"), mean_bearings_plot, width = 12, height = 10)
```

## Comparing Bearings

```{r compare-bearings}

compare_bearings <- function(data1, data2, group1_name, group2_name, bearing_varname) {
    data1 <- data1 %>% rename(bearing = !!bearing_varname)
    data2 <- data2 %>% rename(bearing = !!bearing_varname)

    # Calculate mean bearings
    mean_bearing_1 <- mean.circular(data1$bearing, na.rm = TRUE)
    mean_bearing_2 <- mean.circular(data2$bearing, na.rm = TRUE)

    # Perform Watson-Williams test
    watson_test <- watson.two.test(data1$bearing, data2$bearing)

    # Create tidy table for results
    results_table <- tibble(
        Group = c(group1_name, group2_name),
        Mean_Bearing = c(as.numeric(mean_bearing_1), as.numeric(mean_bearing_2)),
        SD_Bearing = c(sd.circular(data1$bearing, na.rm = TRUE), sd.circular(data2$bearing, na.rm = TRUE))
    )

    # Print results
    print(results_table)
    print(watson_test)

    cb_palette <- c("#D55E00", "#0072B2")

    ggplot() +
        geom_histogram(data = data1, aes(x = bearing, fill = group1_name), alpha = 1, binwidth = 10, boundary = 0) +
        geom_histogram(data = data2, aes(x = bearing, fill = group2_name), alpha = 0.6, binwidth = 10, boundary = 0) +
        coord_polar(start = 0, direction = 1) +
        scale_x_continuous(
            breaks = seq(0, 360, by = 45),
            limits = c(0, 360),
            labels = c("N", "", "E", "", "S", "", "W", "", "N")
        ) +
        scale_fill_manual(values = setNames(cb_palette, c(group1_name, group2_name))) +
        labs(x = NULL, y = "Count", fill = "Bearings") +
        theme(
            panel.grid.major = element_line(color = "gray90"),
            panel.grid.minor = element_blank(),
            legend.position = "bottom"
        ) +
        annotate("segment",
            x = mean_bearing_1, y = 0, xend = mean_bearing_1, yend = Inf,
            arrow = arrow(length = unit(0.5, "cm")), color = cb_palette[1], linewidth = 1
        ) +
        annotate("segment",
            x = mean_bearing_2, y = 0, xend = mean_bearing_2, yend = Inf,
            arrow = arrow(length = unit(0.5, "cm")), color = cb_palette[2], linewidth = 1
        ) +
        annotate("text",
            x = mean_bearing_1, y = Inf,
            label = sprintf("%.1f°", as.numeric(mean_bearing_1)),
            color = cb_palette[1], vjust = 1.1, hjust = -0.1, size = 3
        ) +
        annotate("text",
            x = mean_bearing_2, y = Inf,
            label = sprintf("%.1f°", as.numeric(mean_bearing_2)),
            color = cb_palette[2], vjust = 1.1, hjust = -0.1, size = 3
        )
}

## Compare individual mean bearings
p1 <- compare_bearings(locw_by_id, locf_by_id, "Pups", "Adult Females", "bearing")
p1

p2 <- compare_bearings(locw_by_id, locps_by_id, "Pups", "Particle 0m", "bearing")
p2

p3 <- compare_bearings(locw_by_id, locp_by_id, "Pups", "Particle 186m", "bearing")
p3
```

```{r save-combined-mean-bearings-comparison}
#| echo: false
#| output: false

p2 + p1 + plot_layout(ncol = 2) + plot_annotation(tag_levels = "a")

ggsave(file.path(output_folder, "combined_mean_bearings_comparison.png"), width = 10, height = 6)

p2 + p3 + p1 + plot_layout(ncol = 3) + plot_annotation(tag_levels = "a") &
    theme(text = element_text(size = 10))

ggsave(file.path(output_folder, "combined_mean_bearings_comparison_2.png"), width = 12, height = 6)

```

## Categorise seals as following current if they are within 45 deg of the mean trace direction

```{r categorise-is-following}
#| message: false

### Categorise seals as following current if they are within 45 deg of the mean trace direction
locw_by_id <- locw_by_id %>%
    left_join(locps_by_id %>% st_drop_geometry() %>% select(id, bearing, sd_bearing), by = "id", suffix = c("_seal", "_pt0m")) %>%
    mutate(
        angle_diff_seal_pt0m = angle_diff(bearing_seal, bearing_pt0m),
        is_following_0m = angle_diff_seal_pt0m < 45
    )

# Add 186m particle data
locw_by_id <- locw_by_id %>%
    left_join(locp_by_id %>% st_drop_geometry() %>% select(id, bearing, sd_bearing) %>% rename(bearing_pt186m = bearing, sd_bearing_pt186m = sd_bearing), by = "id") %>%
    mutate(
        angle_diff_seal_pt186m = angle_diff(bearing_seal, bearing_pt186m),
        is_following_186m = angle_diff_seal_pt186m < 45
    )

locw_by_id %>%
    st_drop_geometry() %>%
    left_join(get_survival_data(), by = "id") %>%
    select(id, is_following_0m, is_following_186m, survive_trip_1, survive_year_1) %>%
    pivot_longer(
        cols = c(is_following_0m, is_following_186m),
        names_to = "particle_depth",
        values_to = "is_following"
    ) %>%
    mutate(particle_depth = if_else(particle_depth == "is_following_0m", "0m", "186m")) %>%
    group_by(particle_depth, is_following) %>%
    summarise(
        n = n(),
        prop = (n / nrow(locw_by_id)) %>% round(2),
        n_survived_trip_1 = sum(survive_trip_1, na.rm = TRUE),
        prop_survived_trip_1 = (n_survived_trip_1 / n()) %>% round(2),
        n_survived_year_1 = sum(survive_year_1, na.rm = TRUE),
        prop_survived_year_1 = (n_survived_year_1 / n()) %>% round(2),
        .groups = "drop"
    ) %>%
    mutate(is_following = if_else(is_following, "Following", "Not following")) %>%
    gt(groupname_col = "particle_depth") %>%
    fmt_percent(columns = c(prop, prop_survived_trip_1, prop_survived_year_1), decimals = 0) %>%
    cols_label(
        is_following = "Following Particle",
        n = "Count",
        prop = "Proportion",
        n_survived_trip_1 = "Survived Trip 1",
        prop_survived_trip_1 = "% Survived Trip 1",
        n_survived_year_1 = "Survived Year 1",
        prop_survived_year_1 = "% Survived Year 1"
    ) %>%
    tab_style(
        style = cell_text(weight = "bold"),
        locations = cells_row_groups()
    ) %>%
    cols_merge(
        columns = c(n, prop),
        pattern = "{1} ({2})"
    ) %>%
    cols_merge(
        columns = c(n_survived_trip_1, prop_survived_trip_1),
        pattern = "{1} ({2})"
    ) %>%
    cols_merge(
        columns = c(n_survived_year_1, prop_survived_year_1),
        pattern = "{1} ({2})"
    ) %>%
    cols_label(
        n = "Count (%)",
        n_survived_trip_1 = "Survived Trip 1 (%)",
        n_survived_year_1 = "Survived Year 1 (%)"
    ) %>%
    gtsave(file.path(output_folder, "following_particle_survival.docx"))
```

## Calculate cumulative mean bearings

```{r cumulative-mean-bearings}
#| fig-cap: "Cumulative mean bearings"

# Calculate cumulative mean bearings
calculate_circular_cummean <- function(bearings) {
    n <- length(bearings)

    sapply(1:n, function(i) {
        bearings_subset <- bearings[1:i]
        mean_bearing <- mean.circular(bearings_subset)
    })
}

# Combine weaner and particle data
w_pt <- locw %>%
    left_join(locps %>% st_drop_geometry() %>% select(id, date, bearing), by = c("id", "date"), suffix = c("_seal", "_pt")) %>%
    left_join(locp %>% st_drop_geometry() %>% select(id, date, bearing) %>% rename(bearing_pt186m = bearing), by = c("id", "date")) %>%
    group_by(id) %>%
    mutate(
        cummean_bearing_seal = calculate_circular_cummean(bearing_seal),
        cummean_bearing_pt = calculate_circular_cummean(bearing_pt),
        cummean_bearing_pt186m = calculate_circular_cummean(bearing_pt186m),
        days_since_start = as.numeric(date - first(date)),
        angle_diff_seal_pt = angle_diff(cummean_bearing_seal, cummean_bearing_pt),
        angle_diff_seal_pt186m = angle_diff(cummean_bearing_seal, cummean_bearing_pt186m),
    ) %>%
    left_join(locw_by_id %>% st_drop_geometry() %>% select(id, is_following_0m, is_following_186m), by = "id")



```

## Explore difference in cumulative bearings between seal and particle over time

```{r plot-angle-difference-over-time}

# Plot angle difference over time
p1 <- w_pt %>%
    ggplot(aes(x = days_since_start, y = id)) +
    geom_tile(aes(fill = angle_diff_seal_pt)) +
    scale_fill_viridis_c(option = "B") +
    theme_bw() +
    facet_wrap(~is_following_0m, scales = "free", labeller = as_labeller(c(`TRUE` = "following", `FALSE` = "not following"))) +
    labs(x = "Days since start", y = "Seal ID", fill = "Δ Bearing - 0 m (°)") +
    theme(legend.position = "bottom", text = element_text(size = 8))

p1

p2 <- w_pt %>%
    ggplot(aes(x = days_since_start, y = id)) +
    geom_tile(aes(fill = angle_diff_seal_pt186m)) +
    facet_wrap(~is_following_186m, scales = "free", labeller = as_labeller(c(`TRUE` = "following", `FALSE` = "not following"))) +
    scale_fill_viridis_c(option = "B") +
    labs(x = "Days since start", y = "Seal ID", fill = "Δ Bearing - 200 m (°)") +
    theme(legend.position = "bottom", text = element_text(size = 8))

p2

```

```{r save-angle-difference-tile-plot}
#| echo: false
#| output: false

p1 + p2 + plot_layout(ncol = 2) + plot_annotation(tag_levels = "a") &
    theme(legend.position = "bottom", text = element_text(size = 8))

ggsave(paste0(output_folder, "/angle_diff_tile_plot.png"), width = 10, height = 5, dpi = 300)
```

## Plot angle difference tracks
```{r plot-angle-difference-tracks}
#| fig-cap: "Angle difference tracks"
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 8

## Plot angle difference tracks

plots1 <- w_pt %>%
    rename(x1 = cummean_bearing_seal, x2 = cummean_bearing_pt, y = days_since_start) %>%
    select(id, is_following_0m, x1, x2, y) %>%
    group_split(id) %>%
    map(~ ggplot(data = .x) +
        geom_point(aes(x = x1, y = y, color = "Pups"), size = 0.2, alpha = .5) +
        geom_point(aes(x = x2, y = y, color = "Particle"), size = 0.2, alpha = .5) +
        coord_polar() +
        scale_x_continuous(limits = c(0, 360)) +
        scale_color_manual(values = c("Pups" = "#D55E00", "Particle" = "gray20")) +
        theme_bw() +
        theme(
            legend.position = "bottom", text = element_text(size = 8),
            panel.border = element_rect(color = ifelse(.x$is_following_0m[1], "#0072B2", "grey70"), linewidth = 1)
        ) +
        labs(x = NULL, y = "Days since start", color = "Group", subtitle = .x$id[1]))

o <- order(w_pt %>% group_by(id, is_following_0m) %>% nest() %>% pull(is_following_0m))

plots1 <- plots1[o]

p1 <- wrap_plots(plots1, guides = "collect") +
    plot_annotation(title = "0 m") &
    theme(legend.position = "bottom", axis.title = element_blank())

p1

ggsave(file.path(output_folder, "angle_diff_tracks_0m.png"), p1, width = 10, height = 10, dpi = 300, bg = "white")


plots2 <- w_pt %>%
    rename(x1 = cummean_bearing_seal, x2 = cummean_bearing_pt186m, y = days_since_start) %>%
    select(id, is_following_186m, x1, x2, y) %>%
    group_split(id) %>%
    map(~ ggplot(data = .x) +
        geom_point(aes(x = x1, y = y, color = "Pups"), size = 0.2, alpha = .5) +
        geom_point(aes(x = x2, y = y, color = "Particle"), size = 0.2, alpha = .5) +
        coord_polar() +
        scale_x_continuous(limits = c(0, 360)) +
        scale_color_manual(values = c("Pups" = "#D55E00", "Particle" = "gray20")) +
        theme_bw() +
        theme(
            legend.position = "bottom", text = element_text(size = 8),
            panel.border = element_rect(color = ifelse(.x$is_following_186m[1], "#0072B2", "grey70"), linewidth = 1)
        ) +
        labs(x = NULL, y = "Days since start", color = "Group", subtitle = .x$id[1]))

o <- order(w_pt %>% group_by(id, is_following_186m) %>% nest() %>% pull(is_following_186m))

plots2 <- plots2[o]

p2 <- wrap_plots(plots2, guides = "collect") +
    plot_annotation(title = "200 m") &
    theme(legend.position = "bottom", axis.title = element_blank())

p2

ggsave(file.path(output_folder, "angle_diff_tracks_186m.png"), p2, width = 10, height = 10, dpi = 300, bg = "white")


```

# Survival Model Analysis

```{r survival-model}
#| message: false
#| warning: false

# Prepare data for survival model
all_data_weaners <- load_all_data_weaners() # includes environment data

predictor_vars <- c("sst", "ssha", "eke", "tri", "slope", "SSTgrad", "ice", "dist_to_ice_m", "chl", "chlgrad")

model_data_all <- all_data_weaners %>%
    filter(id %in% w_pt$id, trip == 1, SUS == FALSE) %>%
    group_by(id) %>%
    summarise(across(any_of(predictor_vars), ~ mean(., na.rm = TRUE))) %>%
    left_join(get_survival_data(), by = "id") %>%
    left_join(
        w_pt %>%
            select(id, is_following_0m, is_following_186m) %>%
            st_drop_geometry() %>%
            group_by(id) %>%
            summarise(is_following_0m = last(is_following_0m), is_following_186m = last(is_following_186m)),
        by = "id"
    ) %>%
    select(-contains("seen"))

## Summarise model data
print(model_data_all %>% summary())
print(model_data_all %>% Hmisc::describe())
print(model_data_all %>% psych::describe())

```


```{r}
#| eval: false

## Explore explanatory variables and their relationships with survival
pp <- ggpairs(all_data_weaners %>% select(all_of(c(predictor_vars, "survive_trip_1", "is_following_0m", "is_following_186m"))), aes(color = factor(survive_trip_1)))

ggsave(file.path(output_folder, "model_data_pairs_plot.png"), pp, width = 12, height = 10, scale = 1.5)
```


```{r}
evaluate_model <- function(model) {
    cat("\n\n--- Summary of model_trip (global model) ---\n\n")
    print(model$call)
    tidy(model) %>% print()
    glance(model) %>% print()

    # Add DHARMa model checking for trip survival model
    cat("\n\n--- DHARMa model checking for trip survival model ---\n\n")
    dharma <- simulateResiduals(fittedModel = model, n = 1000)
    plot(dharma)
    testDispersion(dharma) %>% print()
    testZeroInflation(dharma) %>% print()
    testOutliers(dharma) %>% print()

    # model selection
    dredge <- dredge(model)

    best_models <- get.models(dredge, subset = delta <= 2)

    # summarise model selection
    .extract_model_info <- function(model, model_name) {
        glance_info <- glance(model)
        tibble(
            model_name = model_name,
            formula = paste(format(formula(model)), collapse = " "),
            AICc = AICc(model), # Use AICc instead of AIC
            BIC = glance_info$BIC,
            deviance = glance_info$deviance,
            df.residual = glance_info$df.residual,
            null.deviance = glance_info$null.deviance,
            df.null = glance_info$df.null
        )
    }

    model_summary <- map2_dfr(best_models, names(best_models), .extract_model_info) %>%
        mutate(delta_AICc = AICc - min(AICc)) %>% # Use AICc instead of AIC
        arrange(delta_AICc)

    cat("\n\n--- Dredge (model selection) summary ---\n\n")
    print(model_summary %>% as.data.frame())

    # model averaging
    avg_model <- model.avg(best_models)
    cat("\n\n--- Model averaging summary ---\n\n")
    print(summary(avg_model))

    # variable importance
    importance <- sw(dredge)
    cat("\n\n--- Variable importance ---\n\n")
    print(importance)
}
```

## Fit first trip survival model
```{r fit-trip-survival-model}

model_data <- model_data_all %>%
    drop_na(weanmass) %>%
    mutate(across(where(is.numeric), ~ scale(.)))

model_trip <- glm(survive_trip_1 ~ is_following_0m + is_following_186m + weanmass + birthyear + sst + ssha + eke + slope + SSTgrad + ice + chlgrad,
    data = model_data, family = binomial(link = "logit"), na.action = "na.fail"
)

model <- model_trip

evaluate_model(model_trip)

```

## Fit first year survival model
```{r fit-year-survival-model}

## Year survival
model_year <- glm(survive_year_1 ~ is_following_0m + is_following_186m + weanmass + birthyear + sst + ssha + eke + slope + SSTgrad + ice + chlgrad,
    data = model_data, family = binomial(link = "logit"), na.action = "na.fail"
)

evaluate_model(model_year)

```


## Create summary table for all pups
```{r create-summary-table}
#| echo: false
#| output: false

## Create summary table for all weaners
all_data_weaners %>%
    filter(id %in% w_pt$id, trip == 1, SUS == FALSE) %>%
    left_join(get_survival_data()) %>%
    group_by(id, tripdur, birthdate, weanmass, survive_trip_1, survive_year_1) %>%
    summarise(
        max_distance = max(dist2col, na.rm = TRUE) %>% round(0)
    ) %>%
    left_join(
        locw_by_id %>%
            st_drop_geometry(),
        by = "id"
    ) %>%
    ungroup() %>%
    mutate(across(all_of(c("survive_trip_1", "survive_year_1", "is_following_0m", "is_following_186m")), ~ ifelse(. == TRUE, "yes", "no"))) %>%
    mutate(across(where(is.numeric), ~ round(., 1))) %>%
    mutate(across(any_of(c("tripdur", "max_distance")), ~ round(., 0))) %>%
    mutate(
        bearing_seal = paste0(bearing_seal, " ± ", sd_bearing_seal),
        bearing_pt0m = paste0(bearing_pt0m, " ± ", sd_bearing_pt0m),
        bearing_pt186m = paste0(bearing_pt186m, " ± ", sd_bearing_pt186m)
    ) %>%
    select(-c(sd_bearing_seal, sd_bearing_pt0m, sd_bearing_pt186m, angle_diff_seal_pt0m, angle_diff_seal_pt186m)) %>%
    rename(
        "Trip duration (days)" = tripdur,
        "Weaning mass (kg)" = weanmass,
        "Birthdate" = birthdate,
        "Survived trip 1" = survive_trip_1,
        "Survived year 1" = survive_year_1,
        "Max distance to colony (km)" = max_distance,
        "ID" = id,
        "Seal bearing (°)" = bearing_seal,
        "PT bearing 0m (°)" = bearing_pt0m,
        "Followed PT 0m" = is_following_0m,
        "PT bearing 186m (°)" = bearing_pt186m,
        "Followed PT 186m" = is_following_186m,
    ) %>%
    ungroup() %>%
    gt() %>%
    tab_options(
        table.font.size = px(10),
        table.width = px(1000)
    ) %>%
    cols_align(align = "left") %>%
    cols_width(
        starts_with("Birthdate") ~ px(80),
    ) %>%
    sub_missing(missing_text = "") %>%
    gtsave(paste0(output_folder, "/summary_table_weaners.docx"))

```
